Global land cover fraction mapping: methods, variables, challenges

### 2020-05-13 article/article.tex
Writing results section

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{Classification and regression tree}
\makenoidxglossaries
% Disable abbreviating RF for now [text={random forest}]
%\glsunset{RF}

@@

\subsection{Covariate selection}

@@

\begin{enumerate}
 \item RMSE per class comparison: no big differences
 \item Subpixel confusion matrix metrics: quite a big difference
 \item RF variable importance
 \item Truth:prediction scatterplots (hex/box/bar)
 \item Point map of Africa for the different classes
\end{enumerate}

@@

Overall model accuracy statistics are reported in table \ref{tab-accuracy}.
\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.

The models that performed the best (\gls{RF} and Cubist) are related models based on \gls{CART} ensembles.
Predictions per class were made using the binary relevance method, as the models predict a single outcome unlike the \gls{NN} and \gls{MLR} models.
This shows that despite the model training procedure involving less data for each class trained and the rescaling to 100\% requirement, the advantages of these models outweigh the disadvantages of the binary relevance method.
% * <Dainius Masiliūnas> 18:32:47 15 Apr 2020 UTC+0200:
% Discussion point: is a multiple Y RF possible?

All tested models had difficulty discerning the vegetation classes (crops, grassland, shrub, tree) between each other, and less so with the non-vegetated land cover (water, bare soil, urban).

\begin{table}
\centering
\begin{tabular}{llllll}
\toprule
\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & \textbf{\acrshort{NSE}} & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
% * <Dainius Masiliūnas> 14:04:01 13 Aug 2019 UTC+0200:
% Death to kappa
\midrule
Intercept
& 29.9  & 21.4  & 0     & 26±5  & 0.13±0.07 \\
\Gls{PLS} regression
& 27.2  & 17.8  & 0.17  & 38±4  & 0.21±0.06 \\
\Gls{FNC}
& 24.1  & 13.7  & 0.35  & 52±4  & 0.41±0.06 \\
Lasso regression
& 22.1  & 13.3  & 0.45  & 54±4  & 0.40±0.05 \\
\Gls{MLR}
& 21.6  & 12.1  & 0.48  & 58±4  & 0.46±0.06 \\
\Glspl{NN}
% * <Dainius Masiliūnas> 18:06:02 15 Apr 2020 UTC+0200:
% Double-check NN results - were they updated?
& 23.8  & 10.0  & 0.37  & 65±2  & 0.54±0.02 \\
\Gls{SVM} regression
& 20.7  & 8.9   & 0.52  & 69±2  & 0.58±0.03 \\
Cubist regression
& 18.1  & 8.1   & 0.63  & 72±2  & 0.63±0.03 \\
\Gls{RF} regression
& 17.3  & 9.4   & 0.66  & 67±4  & 0.57±0.05 \\
%\ensuremath{''} histogram matched
%& 22.0  & 9.7   & 0.43  & 66±2  & 0.58±0.03 \\
\ensuremath{''} two-step
& 19.9  & 8.2   & 0.56  & 71±2  & 0.62±0.02 \\
\ensuremath{''} three-step
& 19.4  & 8.4   & 0.58  & 71±3  & 0.62±0.04 \\
\ensuremath{''} median vote
& 20.7  & 7.9   & 0.52  & 71±1  & 0.63±0.02 \\
\ensuremath{''} \ensuremath{''} two-step
& 20.0  & 8.1   & 0.54  & 72±1  & 0.63±0.02 \\
\ensuremath{''} \ensuremath{''} three-step
& 20.2  & 7.9   & 0.54  & 72±2  & 0.64±0.02 \\
%\ensuremath{''} \ensuremath{''} histogram matched
%& 21.3  & 9.1   & 0.46  & 68±3  & 0.61      \\
\bottomrule
\end{tabular}
\caption{Accuracy statistics of the models tested.}
\label{tab-accuracy}
\end{table}

% Barplots
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-1m-uncor-bar}
    \caption{Random Forest single model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
    \label{bar-rf-1m-uncor}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-2m-uncor-bar}
    \caption{Random Forest two-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
    \label{bar-rf-2m-uncor}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-3m-uncor-bar}
    \caption{Random Forest three-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
    \label{bar-rf-3m-uncor}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-1m-uncor-hm-bar}
    \caption{Random Forest single model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
    \label{bar-rf-1m-uncor-hm}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-3m-uncor-hm-bar}
    \caption{Random Forest three-step model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
    \label{bar-rf-3m-uncor-hm}
\end{figure}

@@

% 1:1 plots
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-points-a001s2}
    \caption{Random Forest single model prediction correspondence to ground truth (raw points with overplotting).}
    \label{raw-rf-1m-uncor}
\end{figure}
% Binplots
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n100s20}
    \caption{Random Forest single model prediction correspondence to ground truth, 100x100 bins.}
    \label{bin-rf-1m-uncor-n100}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n50s20sqrt}
    \caption{Random Forest single model prediction correspondence to ground truth, 50x50 bins, square root transformation.}
    \label{bin-rf-1m-uncor-n50sqrt}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n50s10sqrt}
    \caption{Random Forest single model prediction correspondence to ground truth, 50x50 bins, square root transformation, smaller circle sizes.}
    \label{bin-rf-1m-uncor-n50sqrts10}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-raster-n50sqrt}
    \caption{Random Forest single model prediction correspondence to ground truth, square root transformation.}
    \label{bin-rf-1m-uncor-rastersqrt}
\end{figure}

% Hexplots
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-1m-uncor-hex}
    \caption{Random Forest single model prediction correspondence to ground truth.}
    \label{hex-rf-1m-uncor}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hex}
    \caption{Random Forest three-step model prediction correspondence to ground truth.}
    \label{hex-rf-3m-uncor}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-1m-uncor-hm-hex}
    \caption{Random Forest single model prediction (histogram matched) correspondence to ground truth.}
    \label{hex-rf-1m-uncor-hm}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hm-hex}
    \caption{Random Forest three-step model prediction (histogram matched) correspondence to ground truth.}
    \label{hex-rf-3m-uncor-hm}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hma-hex}
    \caption{Random Forest three-step model prediction (histogram matched middle, proportionally) correspondence to ground truth.}
    \label{hex-rf-3m-uncor-hma}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hmb-hex}
    \caption{Random Forest three-step model prediction (histogram matched non-extremes) correspondence to ground truth.}
    \label{hex-rf-3m-uncor-hmb}
\end{figure}

% Boxplots
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-1m-uncor-box}
    \caption{Random Forest single model prediction correspondence to ground truth.}
    \label{box-rf-1m-uncor-hm}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-2m-uncor-box}
    \caption{Random Forest two-step model prediction correspondence to ground truth.}
    \label{box-rf-2m-uncor-hm}
\end{figure}
\begin{figure}
    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-3m-uncor-box}
    \caption{Random Forest three-step model prediction correspondence to ground truth.}
    \label{box-rf-3m-uncor-hm}
\end{figure}

>>>>

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\newacronym{OA}{OA}{overall accuracy}
\newacronym{UA}{UA}{user accuracy}
\newacronym{PA}{PA}{producer accuracy}
\newacronym{RMSE}{RMSE}{root mean squared error}
\newacronym{MAE}{MAE}{mean absolute error}
\newacronym{ME}{ME}{mean error}
\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
\newacronym{SDG}{SDG}{Sustainable Development Goal}
\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
\newacronym{SMA}{SMA}{spectral mixture analysis}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{MLP}{MLP}{multi-layer perceptron}
\newacronym{NN}{NN}{neural network}
\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
\newacronym{TPI}{TPI}{Terrain Position Index}
\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
\newacronym{NIR}{NIR}{near infra-red}
\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
\newacronym{FNC}{FNC}{fuzzy nearest centroid}
\newacronym{PLS}{PLS}{partial least squares}
\newacronym{MLR}{MLR}{multinomial logistic regression}
\newacronym{RF}{RF}{random forest}
\newacronym{TOC}{TOC}{Top-of-Canopy}
\newacronym{CART}{CART}{Classification and regression tree}
\newacronym{GLM}{GLM}{General linear regression model}
\makenoidxglossaries
% Disable abbreviating RF for now [text={random forest}]
%\glsunset{RF}

@@

\subsection{Covariate selection}
\label{sec-covariate-selection}

@@

\begin{enumerate}
 \item Differences between all models, showing the RF has lowest RMSE
 \item Variable importance for RF; stressing importance of RS data by a model with only RS variables
 \item Spatial wall-to-wall demo map and spatial errors
\end{enumerate}

@@

Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.

\begin{figure}
    \includegraphics[width=\textwidth]{article-figures/boxplots/2020-05-08-model-comparison-bar}
    \caption{Comparison of the distribution of absolute errors between different models tested, ordered by decreasing overall \gls{MAE}. Dotted lines indicate \gls{MAE}, box indicates median and 75th/25th percentiles.}
    \label{fig-models}
\end{figure}

The models that performed the best (\gls{RF} and Cubist) are related models based on \gls{CART} ensembles.
Predictions per class were made using the binary relevance method (one model per class), as these models are univariate.
This shows that despite the binary relevance limitations of each model only being aware of its relevant class and not the others, and requiring a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models outweigh the disadvantages of the binary relevance method.

Multivariate linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed significantly worse, as they are limited to a linear approach.
The results of all of the linear models were very similar, which shows that there was no added value in \gls{PLS} and Lasso extensions over the basic \gls{GLM}.
This may be due to the additional preprocessing step of covariate selection as detailed in section \ref{sec-covariate-selection}.
It resulted in all remaining covariates being important enough to the prediction of land cover fractions, so that removing any more of them hurts rather than improves model performance.

\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
This indicates that the shape of the covariate data in feature space is too complex to capture merely with a centroid approach.
This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.

\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods, but better than all of the multivariate linear models.

There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large misclassifications or not.
Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk misclassification, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely wrong.

All tested models had difficulty discerning the vegetation classes (grassland, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (water, urban, bare soil).
This was exacerbated by the imbalance in class distribution: the urban class is simple to predict well due to the fact that it is rare and very rarely forms a majority, therefore a guess of 0\% is usually a good one.
In contrast, the tree fraction has a lot of samples with 100\% fraction and it deviates a lot per each sample, making it a highly unpredictable class from the data alone.
Most models showed good improvement in predicting tree cover compared to the intercept model, indicating that the covariates used were useful in telling trees apart from other classes.
In contrast, there were only moderate improvements for predicting the grass fraction, as it is difficult for the classifiers to distinguish this class with the given input covariates.

\begin{table}
\centering
\begin{tabular}{llllll}
\toprule
\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & \textbf{\acrshort{NSE}} & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
% * <Dainius Masiliūnas> 14:04:01 13 Aug 2019 UTC+0200:
% Death to kappa
\midrule
Intercept
& 29.9  & 21.4  & 0     & 26±5  & 0.13±0.07 \\
\Gls{FNC}
& 24.4  & 13.5  & 0.33  & 53±4  & 0.42±0.06 \\
General linear model and \Gls{PLS}
& 21.6  & 12.7  & 0.48  & 56±4  & 0.43±0.05 \\
Lasso regression
& 21.4  & 12.6  & 0.49  & 56±4  & 0.43±0.05 \\
\Gls{MLR}
& 21.6  & 12.1  & 0.48  & 58±4  & 0.46±0.06 \\
Multivariate RF
& 18.7  & 10.0  & 0.61  & 65±3  & 0.55±0.05 \\
\Glspl{NN}
% * <Dainius Masiliūnas> 18:06:02 15 Apr 2020 UTC+0200:
% Double-check NN results - were they updated?
& 22.7  & 9.2   & 0.43  & 68±1  & 0.57±0.02 \\
\Gls{SVM} regression
& 20.7  & 8.9   & 0.52  & 69±2  & 0.58±0.03 \\
Cubist regression
& 18.1  & 8.1   & 0.63  & 72±2  & 0.63±0.03 \\
\Gls{RF} regression
& 17.3  & 9.4   & 0.66  & 67±4  & 0.57±0.05 \\
%\ensuremath{''} histogram matched
%& 22.0  & 9.7   & 0.43  & 66±2  & 0.58±0.03 \\
\ensuremath{''} two-step
& 19.9  & 8.2   & 0.56  & 71±2  & 0.62±0.02 \\
\ensuremath{''} three-step
& 19.4  & 8.4   & 0.58  & 71±3  & 0.62±0.04 \\
\ensuremath{''} median vote
& 20.7  & 7.9   & 0.52  & 71±1  & 0.63±0.02 \\
\ensuremath{''} \ensuremath{''} two-step
& 20.0  & 8.1   & 0.54  & 72±1  & 0.63±0.02 \\
\ensuremath{''} \ensuremath{''} three-step
& 20.2  & 7.9   & 0.54  & 72±2  & 0.64±0.02 \\
%\ensuremath{''} \ensuremath{''} histogram matched
%& 21.3  & 9.1   & 0.46  & 68±3  & 0.61      \\
\bottomrule
\end{tabular}
\caption{Accuracy statistics of the models tested.}
\label{tab-accuracy}
\end{table}

@@
